---
title: "zltrajectory Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zltrajectory Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.height = 5,
  fig.width = 7,
  out.width = "100%"
)
```

## Introduction

`zltrajectory` is an R package designed for analyzing 2D movement data on
zebrafish larvae. Due to their behavior of swimming in discrete bouts,
consisting of an initial burst of speed, this package is primarily designed to
identify where the beginnings of swim bouts occur given the trajectory of a
zebrafish larva.

Use the following commands to download `zltrajectory`.

```{r, eval=FALSE}
require("devtools")
devtools::install_github("RobinEatingWorm/zltrajectory", build_vignettes = TRUE)
```

Use the following command to load and attach the package.

```{r setup}
library(zltrajectory)
```

Use the following command to list all functions available in the package.

```{r, eval=FALSE}
ls("package:zltrajectory")
```

Use the following command to list all sample datasets available in the package.

```{r, eval=FALSE}
data(package = "zltrajectory")
```

## Creating a Trajectory

The `create_trajectory` function builds a data frame that is valid for further
use by other analysis functions within this package. Below, a trajectory is
created from `cadmium_zebrafish`, an example dataset included with this package.

```{r, warning=FALSE}
trajectory <- create_trajectory(cadmium_zebrafish,
                                x = "X.center",
                                y = "Y.center",
                                t = "Trial.time")
head(trajectory, n = 10)
```

The trajectory is formatted as a data frame with three columns. The column `x`
stores the x-position of each observation, the column `y` stores the y-position
of each observation, and the column `t` stores the time of each observation.

## Initial Visualizations

Use the `plot_trajectory` function to visualize the trajectory. Since the
trajectory contains `r nrow(trajectory)` observations, only the 21st minute
(observations with time between 1200 and 1260 seconds) of the dataset will be
visualized.

```{r}
plot_trajectory(trajectory,
                xlab = "x-position (mm)", ylab = "y-position (mm)",
                time_min = 1200, time_max = 1260,
                time_colorbar_lab = "Time\n(s)")
```

The `plot_diagnostics` function creates a plot of distances traveled between
each consecutive pair of observations.

```{r}
plot_diagnostics(trajectory,
                 xlab = "Time (s)", ylab = "Distance (mm)",
                 time_min = 1200, time_max = 1260)
```

To zoom in on the plot, adjust the `time_min` and `time_max` arguments to change
the x-axis plotting limits and the `ylim` argument to change the y-axis plotting
limits.

```{r}
plot_diagnostics(trajectory,
                 xlab = "Time (s)", ylab = "Distance (mm)",
                 time_min = 1240, time_max = 1260, ylim = c(0, 5))
```

It appears that bursts of speed during the zebrafish’s swim bouts can be
detected when it travels over 2 mm between consecutive times recorded.

## Finding Bouts

The `find_bouts` function illustrates the primary functionality of this package.
This function returns a vector of row indices for the trajectory data frame that
correspond to the beginnings of swim bouts.

As shown below, the `min_distance` argument specifies the minimum distance a
fish must travel between consecutive times for a peak (indicating a burst of
speed) in the distance time series to be detected. The `min_time` argument
specifies the minimum time between each detected peak, which is suggested to be
set to 600 ms to avoid cross-talk between swim bouts (Kawashima et al., 2020).

If `return_peaks` is set to `TRUE`, this function returns a list that contains a
vector of row indices for the trajectory that correspond to peaks used to detect
the beginnings of swim bouts in addition to the vector of swim bout row indices
for the trajectory. Each row index with a peak corresponds to one row index with
the beginning of a swim bout and vice versa.

```{r}
bouts_and_peaks_list <- find_bouts(trajectory,
                                   min_distance = 2,
                                   min_time = 0.6,
                                   return_peaks = TRUE)
bouts <- bouts_and_peaks_list$bouts
peaks <- bouts_and_peaks_list$peaks
head(bouts, n = 10)
head(peaks, n = 10)
```

The `plot_diagnostics` function can also be used to assess the correctness of
swim bouts. In the plot showing times between 1240 and 1260 seconds below, the
beginnings of swim bouts are shown in green while the peaks used to detect them
are shown in red. The line in blue shows the minimum distance of 2 required to
detect peaks.

```{r}
plot_diagnostics(trajectory,
                 xlab = "Time (s)", ylab = "Distance (mm)",
                 time_indices = bouts_and_peaks_list, hline = 2,
                 time_min = 1240, time_max = 1260)
```

## Further Analysis

This package includes three additional functions to analyze zebrafish larvae
trajectories.

* `calculate_angles` calculates trajectory angles formed by every consecutive
  group of three observations. If the `bouts` argument is specified, trajectory
  angles formed by every consecutive group of three bouts are calculated
  instead.
* `calculate_bouts_per_time_unit` calculates the number of bouts per unit of
  time specified. The code below calculates the number of bouts per minute since
  the time in the trajectory data frame was recorded in seconds.
* `calculate_distances` calculates distances between every consecutive pair of
  observations. If the `bouts` argument is specified, distances between every
  consecutive pair of bouts are calculated instead.

```{r}
angles <- calculate_angles(trajectory, bouts = bouts)
bouts_per_minute <- calculate_bouts_per_time_unit(trajectory, bouts, unit = 60)
distances <- calculate_distances(trajectory, bouts = bouts)
```

Distributions can be visualized using the `plot_distribution` function. The
following plot shows a histogram of trajectory angles formed by bouts.

```{r}
plot_distribution(angles, "histogram",
                  main = "Distribution of Angles",
                  xlab = "Angle (degrees)",
                  ylab = "Density",
                  breaks = 36,
                  xlim = c(-180, 180))
```

The following plot shows a KDE (kernel density estimate) of the number of bouts
per minute.

```{r}
plot_distribution(bouts_per_minute, "kde",
                  main = "Distribution of Bouts per Minute",
                  xlab = "Bouts per Minute",
                  ylab = "Density")
```

The following plot shows an ECDF (empirical cumulative distribution function) of
distances between bouts.

```{r}
plot_distribution(distances, "ecdf",
                  main = "Distribution of Distances",
                  xlab = "Distance (mm)",
                  ylab = "Cumulative Density")
```

## References

Kawashima, T., Zwart, M. F., Yang, C.-T., Mensh, B. D., & Ahrens, M. B. (2016).
The Serotonergic System Tracks the Outcomes of Actions to Mediate Short-Term
Motor Learning. *Cell*, *167*(4), 933-946.e20.
https://doi.org/10.1016/j.cell.2016.09.055

## Session Information

```{r}
sessionInfo()
```
